---
title: "Clustering on ABC transporter expression"
author: "HJ and NP"
date: "November 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Partitioning around medoids

```{r}
library(FactoMineR)
library(factoextra)
library(mixOmics)
library(cluster)

# load and inspect data
data(multidrug)  # Multidrug data set contains 
dim(multidrug$ABC.trans)  

respca <- PCA(multidrug$ABC.trans, scale = T, graph = F, ncp = 2)
X <- respca$ind$coord
# color by cluster
```

```{r}
fviz_nbclust(respca$ind$coord, pam, method = "silhouette")
```


According to the Silhouette criterion, 3 is a good number of cluster. We will apply a medoid clustering with 3 centers:

```{r}
cl = pam(x = respca$ind$coord, 3)
```


```{r}
fviz_pca_biplot(respca,
axes = c(1,2),
invisible = 'quali',
label = 'var', select.var=list(cos2=0.4),
habillage = as.factor(cl$cluster), pointsize=3, repel=TRUE)

```


```{r}
plot(silhouette(cl))
```


Let's test with an anova if clusters influence the expression of the transporter ABCB5.   
```{r}
lm1 <- lm(multidrug$ABC.trans[,"ABCB5"] ~ as.factor(cl$cluster))
summary(lm1)
```

## Study dataset structure with hierarchical clustering

```{r}
# We remove the column 12 because it contains missing values

X <- multidrug$ABC.trans[,-12]
h1 <- hclust(dist(X), method="single")
h2 <- hclust(dist(X), method="average")
h3 <- hclust(dist(X), method="complete")
plot(h1)
plot(h2)
plot(h3)
```

```{r, fig.width=12}
plot(h3, label = multidrug$cell.line$Sample)
```
We notice that T47D cancer cell line has a very peculiar expression pattern that is not shared by any other cell line.

```{r, fig.width=12}
plot(h3, label = multidrug$cell.line$Class)
```

Some tissue tends to group themselve together (RENAL) but it is far from systematic.

```{r, fig.width=8, fig.height=10}
library(gplots)
library(RColorBrewer)
col_cancer = brewer.pal(9,"Set3")
names(col_cancer) = unique(multidrug$cell.line$Class)
# choose for example the "spectral" palette and create a gradient of colors
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

# We will use euclidean distance and Ward method to build the dendrograms
hclusfun <- function(x) hclust(dist(x), method="ward.D")
row.names(X)=multidrug$cell.line$Class
heatmap.2(X, dendrogram="both", scale="row", 
RowSideColors=col_cancer[multidrug$cell.line$Class], trace="none", col=myPalette, cexRow =1.25,cexCol = 1.25,

hclustfun = hclusfun)
```
