---
title: "PCA in 15 minutes (!)"
author: "EP"
date: "11/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals of Principal Components Analysis 

* Explore the multivariate structure of a dataset
* Decrease dimension of a dataset 
* Visualization tool 
* Explore relationships between variables of a dataset
* Quality control : visualisation of batch effect
* Visualisation of a biological signal (if it exists)

## Data 

* At least 3 quantitative variables (active)
* A few categorical variables (supplementary)
* Don't forget to transform your data if needed (try log)

## Principle of the method 

### With maths

* Computation of eigen vectors and eigen values of the correlation matrix between variables
* Projection into the space of eigen vectors
* Iterated power algorithm

### With hands

* Projection of the data into a subspace that maximizes the distances among observations 
* Maximization of the inertia 

## Example with R on iris dataset

What you need to tune when performing PCA 

* Number of axes to plot
* Centering and scaling? 
* For variables : explore correlations : define a threshold on contribution
* Coloration by qualitative variables

```{r}
library(FactoMineR)
library(factoextra)

# Load the iris dataset
data(iris)
# read the help
?iris

# compute the PCA decomposition
respcairis <- PCA(iris[,1:4], scale = T, graph = F)
# check the percentage of variance captured by each component
respcairis
respcairis$eig
fviz_screeplot(respcairis, addlabels = TRUE)

# Plot the samples in the space spanned by PC1 and PC2
# color by species
fviz_pca_ind(respcairis,
axes = c(1,2),
invisible = 'quali',
label = 'none',
habillage = iris$Species)

# color by cos2
fviz_pca_ind(respcairis,
axes = c(1,2),
invisible = 'quali',
label = 'none',
col.ind = 'cos2')

# Plot the variable contributions to the PCs
fviz_pca_var(respcairis,
axes = c(1,2))
library("corrplot")
corrplot(respcairis$var$cos2)

```

